---
title: "Thesis Ch. 3 Code"
author: "Hannah Girgente"
date: "2024-03-31"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r Working Directory}
setwd("E:/McIntyre_Lab/Thesis/Data/OdonateCommunity")
```

## Packages used:
```{r packages, warning=FALSE}
library(BiodiversityR)
library(labdsv)
library(MASS)
library(MVA)
library(vegan)
library(plyr)
library(betapart)
library(ggplot2)
library(ggrepel)
library(grid)
library(corrplot)
library(lattice)
library(picante)
library(stats)
library(sars)
library(dplyr)
library(optpart)
library(cluster)
```

## Data used
```{r data}
#Data that includes adults seen at each site
odes <- read.csv("odes-withadult.csv", row.names = 1, header = T)
View(odes) 

#Data for only the nymphs sampled at each site
odes.comm <- read.csv("nmds_odes.csv", row.names = 1, header = T)
View(odes.comm)

ode.matrix <- as.matrix(odes.comm) #convert the nymph dataset to a matrix

#All site types (steephead or not)
sites <- read.csv("oa_sites.csv", row.names = 1, header = T)

#Sites types, exclusing sites that had no nymphs sampled at them
sites.nmds <- read.csv("oa_sites-nmds.csv", row.names = 1, header = T)
View(sites.nmds)

#Environmental data for each site
env <- read.csv("nmds_env.t.csv", row.names = 1, header = T)
env_scale <- scale(env, center = F, scale = T) #scale the data, but not centered to avoid negatives
env_matrix <- as.matrix(env_scale) #convert the data frame into a matrix
View(env_matrix)

#Environmental data (without the stream comp) for correlation analyses
corr.env <- read.csv("corr_env.trans.csv", row.names = 1, header = T)
corr.env.s <- scale(corr.env, center = F, scale = T) #scale the data
View(corr.env.s)

```

# Species richness and diversity indicies
```{r diversity}
#Species richness per site
ode.rich <- specnumber(odes)
ode.rich

#Richness estimate
sp_est <- specpool(odes) 
sp_est

#Examine the richness across steephead and non-steephead sites
ode.sp <- specpool(odes, pool = sites$type)
ode.sp

#Richness and estimates for steephead sites
ode.sp["Steephead",]

#Richness and estimates for non-steephead sites
ode.sp["Non-Steephead",]

#If you want a particular richness metric for all stream types
ode.sp[,c("boot", "boot.se")]

#species accumulation curves for estimates of total species richness based on 
#iterative sampling with confidence intervals:
(ode.sp = poolaccum(odes))
plot(ode.sp)

#To plot just one of these metrics on their own:
ode.est <- poolaccum(odes)
ode.est
plot(ode.est, col="black", 
     strip=function(..., bg) strip.default(..., bg="gray90"), 
     display=c("boot", "S"))

#Which stream type was the most diverse
#Plot the average number of species at each steephead vs non-steephead site
boxplot(specnumber(odes) ~ type, data = sites, xlab= "Stream type", 
        ylab = "# of species")

#Calculate the shannon's diversity index for each site
diversity(odes[-1], index = "shannon")

sites.2 <- read.csv("oa_sites.csv", row.names = 1, stringsAsFactors = T)

#Shannon's diversity index for each stream type
diversitycomp(odes, y = sites.2, factor1 = "type", index ="Shannon",
              method = "pooled")

#Which stream type had individuals the most evenly distributed over space
diversitycomp(odes, y=sites.2, factor1="type", index="Jevenness", 
              method = "pooled")

#Which stream type had the highest alpha diversity
alpha <- tapply(specnumber(odes), sites, FUN= mean)
alpha

#How similar were steephead sites to non-steephead sites (beta diversity)
ode <- rowsum(odes, group= sites$type)
betadiver(ode, method="w")



#Is beta diversity significantly different between steepheads and non-steepheads?
ode.b <- betadiver(odes, method = "w")
ode.ano <- anosim(ode.b, sites$type)
ode.ano


#Check if any sites are outliers with respect to ode richness
occupancy <- read.csv("nmds_odes.csv", header=TRUE)
boxplot(specnumber(occupancy), xlab = "Sites 1-23", ylab = "# of species")

ano <- anosim(ode.matrix, sites.nmds$type, distance = "bray", permutations = 9999)
ano

#species occurrence patterns
abuocc(odes)
```

# Mantel Tests
### (test to see if any environmental variables have a statistically significant impact on odonate assemblage structure)
```{r mantel tests}

### Canopy ###
canopy <- env_matrix[,1]
View(canopy)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.can <- dist(canopy, method = "euclidean")
ode.can <- mantel(dist.ode, dist.can, method = "spearman", permutations = 9999, na.rm = T)
ode.can
### p= 0.0561

### Temperature ###
temp <- env_matrix[,2]
View(temp)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.temp <- dist(temp, method = "euclidean")
ode.temp <- mantel(dist.ode, dist.temp, method = "spearman", permutations = 9999, na.rm = T)
ode.temp

### Velocity ###
v <- env_matrix[,3]
View(v)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.v <- dist(v, method = "euclidean")
ode.v <- mantel(dist.ode, dist.v, method = "spearman", permutations = 9999, na.rm = T)
ode.v

### pH ###
ph <- env_matrix[,4]
View(ph)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.ph <- dist(ph, method = "euclidean")
ode.ph <- mantel(dist.ode, dist.ph, method = "spearman", permutations = 9999, na.rm = T)
ode.ph

### DO ###
do <- env_matrix[,5]
View(do)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.do <- dist(do, method = "euclidean")
ode.do <- mantel(dist.ode, dist.do, method = "spearman", permutations = 9999, na.rm = T)
ode.do

### TDS ###
tds <- env_matrix[,6]
View(tds)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.tds <- dist(tds, method = "euclidean")
ode.tds <- mantel(dist.ode, dist.tds, method = "spearman", permutations = 9999, na.rm = T)
ode.tds

### Width ###
wid <- env_matrix[,7]
View(wid)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.wid <- dist(wid, method = "euclidean")
ode.wid <- mantel(dist.ode, dist.wid, method = "spearman", permutations = 9999, na.rm = T)
ode.wid
# p=0.0566

### Turbidity ###
turb <- env_matrix[,8]
View(turb)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.turb <- dist(turb, method = "euclidean")
ode.turb <- mantel(dist.ode, dist.turb, method = "spearman", permutations = 9999, na.rm = T)
ode.turb

### Depth ###
dep <- env_matrix[,9]
View(dep)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.dep <- dist(dep, method = "euclidean")
ode.dep <- mantel(dist.ode, dist.dep, method = "spearman", permutations = 9999, na.rm = T)
ode.dep

### Sand ###
sand <- env_matrix[,10]
View(sand)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.sand <- dist(sand, method = "euclidean")
ode.sand <- mantel(dist.ode, dist.sand, method = "spearman", permutations = 9999, na.rm = T)
ode.sand

### Gravel ###
grv <- env_matrix[,11]
View(grv)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.grv <- dist(grv, method = "euclidean")
ode.grv <- mantel(dist.ode, dist.grv, method = "spearman", permutations = 9999, na.rm = T)
ode.grv

### Cobble ###
cob <- env_matrix[,12]
View(cob)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.cob <- dist(cob, method = "euclidean")
ode.cob <- mantel(dist.ode, dist.cob, method = "spearman", permutations = 9999, na.rm = T)
ode.cob

### Boulders ###
bo <- env_matrix[,13]
View(bo)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.bo <- dist(bo, method = "euclidean")
ode.bo <- mantel(dist.ode, dist.bo, method = "spearman", permutations = 9999, na.rm = T)
ode.bo

### fpom ###
fpom <- env_matrix[,14]
View(fpom)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.fpom <- dist(fpom, method = "euclidean")
ode.fpom <- mantel(dist.ode, dist.fpom, method = "spearman", permutations = 9999, na.rm = T)
ode.fpom

### cpom ###
cpom <- env_matrix[,15]
View(cpom)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.cpom <- dist(cpom, method = "euclidean")
ode.cpom <- mantel(dist.ode, dist.cpom, method = "spearman", permutations = 9999, na.rm = T)
ode.cpom

### Bryophytes ###
bry <- env_matrix[,16]
View(bry)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.bry <- dist(bry, method = "euclidean")
ode.bry <- mantel(dist.ode, dist.bry, method = "spearman", permutations = 9999, na.rm = T)
ode.bry

### Small wood ###
sw <- env_matrix[,17]
View(sw)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.sw <- dist(sw, method = "euclidean")
ode.sw <- mantel(dist.ode, dist.sw, method = "spearman", permutations = 9999, na.rm = T)
ode.sw

### Large wood ###
lw <- env_matrix[,18]
View(lw)
dist.ode <- vegdist(ode.matrix, method = "bray", binary = T)
dist.lw <- dist(lw, method = "euclidean")
ode.lw <- mantel(dist.ode, dist.lw, method = "spearman", permutations = 9999, na.rm = T)
ode.lw

### Cumulative effect of all environmental variables ###
dist.env <- dist(env_matrix, method = "euclidean")
ode.env <- mantel(dist.ode, dist.env, method = "spearman", permutations = 9999, na.rm = T)
ode.env

```

# Non-metric Multidimensional Scaling (NMDS) ordination
```{r NMDS}
set.seed(123) #make the result repeatable

#run the NMDS ordination for binary data using the matrix you made in the matel test section
meta.ode <- metaMDS(ode.matrix, distance = "bray", k= 3, 
                    binary = T, weakties = F, autotransform = F)
meta.ode
plot(meta.ode)
#extract the data scores into a data frame
data.scores <- as.data.frame(scores(meta.ode)$sites)
#add the stream type to the data frame for each site
data.scores$type <- sites.nmds$type

#see which # of dimensions results in the largest descrease in stress
for(i in 2:5) print(metaMDS(ode.matrix, distance = "bray", k=i,
                            binary = T, weakties = F, autotransform = F, trace = F)$stress*100)
#3 dimensions is best

#Plot the stress values by # of dimensions
stressplot(meta.ode)

## Plot the site x species NMDS in ggplot ##
nmds.plot <- ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(size = 4, aes(shape = type, color = type)) +     # make point shape/color correlate to stream type
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
        axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
        legend.text = element_text(size = 12, face ="bold", colour ="black"), 
        legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
        axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
        legend.title = element_text(size = 14, colour = "black", face = "bold"), 
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
        legend.key=element_blank()) + 
  labs(x = "NMDS1", color = "Type", y = "NMDS2", shape = "Type")  +     #Add a legend 
  scale_colour_manual(values = c("#009E73", "#E69F00"))

nmds.plot


## envfit to add environmental data to the NMDS plot ##

#Correlation analyses to decide if any metrics should be removed
env_cor <- corr.env
env_cor <- as.matrix(corr.env)
env_cor <- na.omit(env_cor) #remove rows with empty values
cor_mat <- cor(env_cor, method = 'pearson') #use pearson method to construct correlation matrix
corrplot(cor_mat, method = "number", type = "lower", diag = F, tl.col = "black") #plot the lower half of the correlation matrix
#remove spc due to high correlation with tds

#create environmental vectors for each variable to fit onto NMDS plot
en <- envfit(meta.ode, env_matrix, permutations = 999, na.rm = T) #use env_matrix object from the mantel tests earlier
en

#make the length of vector related to strength of correlation
en_coord <- as.data.frame(scores(en, "vectors")) * ordiArrowMul(en) 

#Plot the nmds with the environmental data fitted to it
nmds.env <- ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(data = data.scores, size = 4, aes(shape = type, color = type)) +
  geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2),  # Add the environmental vectors as line segments
               data = en_coord, linewidth =1, alpha = 0.5, color = "grey30") +
  geom_text_repel(data = en_coord, aes(x = NMDS1, y = NMDS2), color = "grey30", 
                  fontface = "bold", label = row.names(en_coord)) +
  annotate(geom = "label", x = -1, y = 1.25, size = 5,
           label = paste("Stress: ", round(meta.ode$stress, digits = 3))) +  #Add the stress value to the plot
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
        axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
        legend.text = element_text(size = 12, face ="bold", colour ="black"), 
        legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
        axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
        legend.title = element_text(size = 14, colour = "black", face = "bold"), 
        panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.2),
        legend.key=element_blank()) + 
  labs(x = "NMDS1", color = "Type", y = "NMDS2", shape = "Type")  +     #Legend
  scale_colour_manual(values = c("#009E73", "#E69F00"))
nmds.env
```

# Cluster analysis
### (see how sites group together by assemblage structure and by environmental data)
```{r cluster}
### Environmental data ###
env_clust <- hclust(dist.env, method = "average")
plot(env_clust, xlab = "Site by environment", ylab = "Euclidean dissimilarity", hang = -2)

#See how correlated dissimilarity matrix is with the conphenetic dist. 
#(how well does our dendrogram fit our data)
env.conph <- cophenetic(env_clust)
cor(dist.env, env.conph)

# Silhouette plot
envpam <- pam(dist.env, k=2)
attributes(envpam)

envpam$medoids
envpam$clustering
plot(envpam)

clusplot(envpam, labels = 2, cex = 0.65)

### Odonate data ###
#Create a dissimilarity matrix for the odonate data at each site using bray-curtis
clust.dist <- vegdist(odes, method = "bray", binary = T)

#Make a hierarchical cluster dendrogram, using average linkage
odes.clust <- hclust(clust.dist, method = "average")
plot(odes.clust, xlab = "Site by species", ylab = "Bray-Curtis' dissimilarity", hang = -2) #use hang to bring the site names in line with each other

#Calculate the distance at which each each object merges with another on the dendrogram
comm.coph <- cophenetic(odes.clust)

#See how correlated dissimilarity matrix is with the conphenetic dist. 
#(how well does our dendrogram fit our data)
cor(clust.dist, comm.coph) #pretty good fit

odepam <- pam(clust.dist, k=3)
attributes(odepam)

odepam$medoids
odepam$clustering
plot(odepam)

clusplot(odepam, labels = 3, cex = 0.65)
```

# Stacked species richness plot
### separated by life stage (adult and nymphs documented at each site)
```{r stacked plot}
#Create data frame for species richness separated by adult and nymph species at each site
#Make an object containing all the sites sampled
sites.graph <- c("RES01", "RES01", "RES03", "RES03", "RES04", "RES04","RES05",
                 "RES05", "WB01", "WB01", "TRSF01", "TRSF01", "BB01", "BB01", 
                 "CCR01", "CCR01","LCWMA01", "LCWMA01", "LCWMA02","LCWMA02", 
                 "LCWMA03","LCWMA03", "JB01", "JB01","PB01","PB01", "EAFB01", 
                 "EAFB01", "EAFB02", "EAFB02", "EAFB03","EAFB03","BRSF01", 
                 "BRSF01", "BC01", "BC01", "LTSF01","LTSF01", "LTSF02", "LTSF02",
                 "ARBP01", "ARBP01","TSP01", "TSP01", "TSP02", "TSP02","CB01", "CB01")
# repeat each site twice, once for nymphal observations and another for adult
sites.graph

#Create a data frame with sites on x and its associated adult/nymphal species richness on y
ode.comm <- data.frame(Site = sites.graph, sp.rich = c(3,0,3,0,5,3,2,0,8,0,5,1,2,
                                                       0,2,0,6,3,1,4,10,4,5,4,1,
                                                       4,8,5,4,0,0,3,7,0,6,1,12,
                                                       0,3,2,11,11,5,3,1,1,9,3))
#Add a column to the data frame for life stage for each value of richness at a site
ode.comm$stage <- c("Nymph", "Adult", "Nymph", "Adult", "Nymph", "Adult", 
                    "Nymph", "Adult", "Nymph", "Adult", "Nymph", "Adult", 
                    "Nymph", "Adult", "Nymph", "Adult", "Nymph", "Adult",  
                    "Nymph", "Adult", "Nymph", "Adult", "Nymph", "Adult", 
                    "Nymph", "Adult", "Nymph", "Adult", "Nymph", "Adult",
                    "Nymph", "Adult", "Nymph", "Adult","Nymph", "Adult",
                    "Nymph", "Adult","Nymph", "Adult","Nymph", "Adult",
                    "Nymph", "Adult","Nymph", "Adult","Nymph", "Adult")
# alternate adult nymph so that each site has an adult and a nymph column
ode.comm

#Make Site a factor and arrange the sites so that all non-steepheads are together 
#on the left and all steepheads are together on the right
ode.comm$Site <- factor(ode.comm$Site, levels = c("RES01", "RES03", "RES04",
                                                  "RES05","WB01", "TRSF01", 
                                                  "BB01", "CCR01","LCWMA01", 
                                                  "LCWMA02","LCWMA03","JB01",
                                                  "PB01","EAFB01", "EAFB02", 
                                                  "EAFB03","BRSF01","BC01", 
                                                  "LTSF01", "LTSF02","ARBP01",
                                                  "TSP01","TSP02","CB01"), ordered = T)

# Sort by site and life stage (adult or nymph)
odecomm_sort <- arrange(ode.comm, Site, stage)
head(odecomm_sort)

# Calculate the cumulative sum of sp.rich for each site
ode_cumsum <- ddply(odecomm_sort, "Site", transform, tot.rich=cumsum(sp.rich))

# change the data source in ggplot to this new data frame

#Stacked bar graph of species richness per site broken down to number of species
#documented as nymphs and number of species documented as adults
ode.plot <- ggplot(data = ode_cumsum, aes(x=Site, y=sp.rich, label=sp.rich, fill=stage)) + #color the two parts of the bar by life stage
  geom_bar(stat = "identity") + #display the values in the data frame as is
  geom_text(data=subset(ode.comm, sp.rich != 0), aes(y=sp.rich), 
            position = position_stack(vjust = 0.5), color="white", size=3.5) + #Add the sp. rich values to the bars
  scale_fill_brewer(palette = "Paired") +
  labs(x = "Site", y = "Species richness", fill = "Life stage") + #Axis and legend labels
  theme_classic() +
  theme(panel.grid= element_blank(), 
        axis.title = element_text(size = 16), 
        legend.text = element_text(size = 16), 
        legend.title = element_text(size = 16),
        axis.text = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10,
                                 color = "black")) + #Make the axis labels vertical rather than horizontal 
  scale_y_continuous(name = "Species richness", limits = c(0,25)) #Set the upper and lower bounds of the y axis
ode.plot

```


